Adaptive evolution with aneuploidy and mutation

### 2021-11-08 ms/ms.tex
added config description for pyabc

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC}\footnote{https://pyabc.readthedocs.io} Python 
package~\citep{Klinger2018}.
Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed. 
After getting weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels. $50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
% TODO ILIA please add all configuration details, similar to the relevant section in the NPI paper. Discuss priors, convergence tests, etc.
% And also number of replicates for different models

@@

For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-8,1e-5)$ and $\mathit{Uniform}(1e-6,1e-3)$ respectively. We have tried also uniform priors with lower than $1e-8$ bound for $\mu$, but anyway the posterior ends with higher values. 
For the single locus model without aneuploidy, for the mutation rate wider uniform prior was used - $\mathit{Uniform}(1e-10,1e-5)$. 
%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
%TODO ILIA describe the priors for the mutation and aneuploidy rate.
% TODO also describe the prior for the model without aneuploidy - the mutation rate is wider here.

>>>>

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC}\footnote{https://pyabc.readthedocs.io} Python 
package~\citep{Klinger2018}.
Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed. 
%adapted description from Syga et al. 2021:
In short, the algorithm employs sequential importance sampling over generations. In generation $t$, the algorithm draws sets of parameters $\theta$ from a given proposal distribution and consequently simulates data $C$ from the model until $n_t$ instances were accepted ($n_t$ "particles") based on the comparison to observed data $C_{obs}$ via a distance function $D(C, C_obs)$ and acceptance threshold $\epsilon_t$, such that $D(C, C_objs)\leq{\epsilon_t}$. As the distance function we chose $1-\likelihood$. For the acceptance threshold the median of the distances of the accepted instances of the previous generation were applied (for the first generation the acceptance threshold is defined $0.01$). 
When $n_t$ instances were accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix. In the first generation the proposal distribution equals to the prior. 
The size of $n_t$ was determined by adaptive population strategy~\citep{Klinger2017}.
We stop the iterations, when no more sufficient change in acceptance threshold occur, and no change in posterior.
After receiving weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels. $50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
% TODO convergence tests, etc. Number of replicates for different models.

@@

For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-8,1e-5)$ and $\mathit{Uniform}(1e-6,1e-3)$ respectively. We have tried also uniform priors with lower than $1e-8$ bound for $\mu$, but anyway the posterior ends with higher values. 
For the single locus model without aneuploidy, for the mutation rate wider uniform prior was used - $\mathit{Uniform}(1e-10,1e-5)$. 
%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
